\textbf{Dynamic Hover Detection:\\Creating GUIs in Elm}

### 2013-12-15 paper/paper.tex
Redid section 4 (explaining our implementation of menus). It flows a lot better now.

Using \texttt{hoverablesJoin}, we created a structure of type
\texttt{Tree (Signal Element, Signal Bool)} to represent our menus.
Using the combinator
\texttt{combine : {[}Signal a{]} -\textgreater{} Signal {[}a{]}}, we
were able to turn this structure into a
\texttt{Signal (Tree (Element, Bool))}. Once we had that structure, we
could render the menu using a pure function from
\texttt{Tree (Element, Bool) -\textgreater{} Element} and then lift it.
Similarly to limiting the use of the IO monad in Haskell code, it is
advantageous to limit the use of signals in Elm code.

Our first trick was to replace submenus not shown on the screen with
spacers. A spacer is simply a blank rectangle with a specified with and
height
(\texttt{spacer : Int -\textgreater{} Int -\textgreater{} Element}). For
every menu item, we create both the menu Element and a spacer. The
spacer has the same width as the Element, but a height of 1. We then
switch between menu and the spacer based on hover information. This
helped in two ways. First of all, it allowed us to choose between two
options without creating a signal of signals. Secondly, the spacer
helped to align the submenu with its parent element.

Our second trick fixed a race condition. A submenu was rendered to the
screen if one of three conditions was met: the mouse was hovering upon
its parent, the mouse was hovering upon it, or the mouse was hovering
upon any of its descendants. As the mouse moved from the parent to the
child, the child was often replaced with a spacer before it could detect
that the mouse was hovering over it. To fix this problem, we created the
following function:

@@

Using \texttt{extend} makes a \texttt{Signal Bool} wait a millisecond to
transition from True to False. By applying this function to every
hovering Boolean in the Menu structure, we gave the mouse time to move
from a menu Element to its submenu before that submenu disappeared.

@@

Note that the input to our menu is a \texttt{Tree (Signal String)}, but
not a \texttt{Signal (Tree String)}. We are able to map from
\texttt{Signal Element -\textgreater{} (Signal Element, Signal Bool)},
but not from \texttt{Element -\textgreater{} (Element, Bool)}.
Therefore, if we tried to turn a \texttt{Signal (Tree String)} into a
\texttt{Signal (Tree (Element, Bool))}, we would need to lift an
impossible hoverable function of type
\texttt{Element -\textgreater{} (Element, Bool)} on each element.
Therefore, each individual menu element can contain dynamic information,
but the menu structure must be static. We can, in practice, get around
this restriction by creating a tree that is larger than necessary and
filling the unused nodes with empty string, which our implementation
handles appropriately.

>>>>

Using \texttt{plainText : String -> Element}, we can take a menu specification
of type \texttt{Tree (Signal String)} and turn into a \texttt{Tree (Signal
Element}. From there, we can use \texttt{hoverablesJoin} to get the hover
information of each element and create a structure of type \texttt{Tree (Signal
Element, Signal Bool)} to represent our menus.  By recursing on the combinator
\texttt{combine : {[}Signal a{]} -\textgreater{} Signal {[}a{]}}, we are able to
turn this structure into a \texttt{Signal (Tree (Element, Bool))}. By changing
our internal structure into a signal of trees instead of a tree of signals, we
can do the rest of our work with pure functions that operate on the tree, rather
than on individual signals.  For example, our rendering function
\texttt{renderMenu : Tree (Element, Bool) -\textgreater{} Element} can simply be
lifted onto the menu structure.  Similarly to limiting the use of the IO monad
in Haskell code, it is advantageous to limit the use of signals in Elm code.

The first problem in implementing menus is twofold: how to position submenus on the
screen and how to hide submenus that should not be rendered. To get around these
problems, we replace submenus not shown on the screen with
spacers. A spacer is simply a blank rectangle with a specified width and
height
(\texttt{spacer : Int -\textgreater{} Int -\textgreater{} Element}). For
every menu item, we create both the menu Element and a spacer. The
spacer has the same width as the Element, but a height of 1. We then
switch between menu and the spacer based on hover information. This
helped in two ways. First of all, it allowed us to choose between two
options without creating a signal of signals. Secondly, the spacer
helped to align the submenu with its parent element.

\begin{verbatim}
elemOrSpacer : (Element, Bool) -> Element
elemOrSpacer (elem, hover) =
    if hover
    then elem
    else spacer (widthOf elem) 1
\end{verbatim}

By replacing hidden elements with spacers, we are able to use the \texttt{flow}
primitive to position our elements instead of manually doing ``absolute''
positioning.  A submenu is rendered to the screen if one of three conditions
is met: the mouse is hovering upon its parent, the mouse is hovering upon it,
or the mouse is hovering upon any of its descendants.

While this allows recursive submenus to be correctly rendered to the screen, it
creates a race condition. As the mouse moves from the parent to the its submenu,
the child is often replaced with a spacer before it can detect that the mouse is
hovering over it. The results in menus disappearing as the mouse changes which
element it is hovering upon. To fix this problem, we created the following
function:

@@

\texttt{extend} makes a \texttt{Signal Bool} wait a millisecond to
transition from True to False. By applying this function to every
hovering Boolean in the Menu structure, we give the mouse time to move
from a menu Element to its submenu before that submenu disappears.

A menu that does not detect mouse clicks is a fairly useless menu. There is no
primitive in Elm that can simply attach click detection to an element. The
closest primitive is \texttt{customButtons}. However, the implementation of
\texttt{customButtons} creates an extra dummy spacer which is rendered on top of
the element and therefore prevents us from changing the color of the menu
element when the mouse is hovering over it. To get around this issue, we used a
combination of \texttt{Mouse.clicks : Signal ()} (which fires an event whenever
the mouse clicks anywhere on the screen) and the hoverable information to
determine which element (if any) the mouse was hovering over.

\begin{verbatim}
clicksFromHover : Signal Bool -> Signal ()
clicksFromHover hover =
    lift (\_ -> ()) <| keepIf id False <| sampleOn Mouse.clicks hover
\end{verbatim}

For each menu element in the tree, we found its ``path'' (with type
\texttt{[Int]}) which denotes the list indexes needed to traverse to that
element. Combining this with the click information gave each element a signal of
its path that fired whenever the user clicked on that element.  We then combined
all of those signals into one signal using the \texttt{merge} combinator
(\texttt{merges : [Signal a] -> Signal a}). The user of the menu could then use
this \texttt{Signal [Int]} to respond to mouse clicks. This is significantly
more flexible than returning a \texttt{Signal String} because a path is
guaranteed to be unique, whereas a menu string might have a duplicate.

@@

Also note that the input to our menu is a \texttt{Tree (Signal String)}, but
not a \texttt{Signal (Tree String)}. We are able to map from
\texttt{Signal Element -\textgreater{} (Signal Element, Signal Bool)},
but not from \texttt{Element -\textgreater{} (Element, Bool)}.
Therefore, if we tried to turn a \texttt{Signal (Tree String)} into a
\texttt{Signal (Tree (Element, Bool))}, we would need to lift an
impossible hoverable function of type
\texttt{Element -\textgreater{} (Element, Bool)} on each element.
Therefore, each individual menu element can contain dynamic information,
but the menu structure must be static. We can, in practice, get around
this restriction by creating a tree that is larger than necessary and
filling the unused nodes with empty string, which our implementation
handles appropriately.
